---
title: "Why (R) Markdown?"
author: "Andrew Whitby"
date: "4 April 2017"
output: ioslides_presentation
runtime: shiny
---

## Authoring process

### Traditional

1. Download data from databank
2. Do analysis in Excel
3. Draft charts in Excel
4. Write text in Word
5. Style (& probably remake figures)

## Authoring process

### R Markdown

1. Access data from API if possible
2. Write analysis code in R
3. Output charts in R (ggplot)
4. Write text in Markdown
5. Style

## What is markdown?

Markdown (a play on 'markup')

## Interactive bit

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```


